---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rDataSet

<!-- badges: start -->
<!-- badges: end -->

The goal of rDataSet is to treat data frames as mathematical sets. It provides set-theoretic operations (union, intersection, difference, equality) for data frames where values are identified by row IDs and column names, and presence/absence is determined by `NA` values.

## Installation

You can install the development version of rDataSet from local source like so:

``` r
devtools::install("/path/to/rDataSet")
```

## Overview

rDataSet introduces a `dataset` S3 class that extends data frames with designated ID columns. Set operations work on a cell-by-cell basis:

- **ID columns** uniquely identify each row (one-to-one relation)
- **Value presence** is determined by `is.na()` - a value is present if not `NA`
- **Set operations** compare cells at the same (row_id, col_id) position

## Set Operations

| Operation | Syntax | Description |
|-----------|--------|-------------|
| Difference | `A - B` | Values in A where corresponding cells in B are `NA` |
| Intersection | `A > B` | Values in A where corresponding cells in B are not `NA` |
| Union | `A + B` | Values from A, falling back to B where A has `NA` |
| Equality | `A == B` | Cell-wise comparison returning logical values |

## Example

```{r example}
library(rDataSet)
library(tibble)
library(dplyr)

# Create two datasets with 'i' as the ID column
A <- dataset_build(
  tibble(i = 1:10, b = if_else(1:10 %% 2 == 0, NA, 1:10)),
  ids = "i"
)

B <- dataset_build(
  tibble(i = 1:10, b = na_if(1:10, 3)),
  ids = "i"
)

# View the datasets
A
B
```

### Set Difference

Keep values from A where B has `NA`:

```{r difference}
A - B
```

### Set Intersection

Keep values from A where B has non-`NA` values (B acts as a filter):

```{r intersection}
A > B
```

### Set Union

Combine values from both datasets, with A taking precedence:

```{r union}
A + B
```

### Set Equality

Compare datasets cell-by-cell:

```{r equality}
A == A  # All TRUE (or NA for missing values)
A == B  # Shows which cells match
```

## Helper Functions

```{r helpers}
# Extract ID columns
ids(A)

# Extract value columns (non-ID columns)
vals(A)
```

## Mathematical Properties

The set operations satisfy the fundamental identity: `a = (a ∩ b) ∪ (a \ b)`

```{r identity}
# Verify: A == (A > B) + (A - B)
all(vals(A == (A > B) + (A - B)), na.rm = TRUE)
```

## License

GPL (>= 3)
