---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rDataSet

<!-- badges: start -->
<!-- badges: end -->

The goal of rDataSet is to treat data frames as mathematical sets. It provides set-theoretic operations (union, intersection, difference, equality) for data frames where values are identified by row IDs and column names, and presence/absence is determined by `NA` values.

On top of Operations, Datasets support writing and reading to json, toml files. With the goal of having humand readable
Data represntations which don't relay on table represnataions. (as there is a lack of decent diffing and viewing for such data, at least thats how i perceive it)

## Installation

You can install the development version of rDataSet from local source like so:

``` r
devtools::install("/path/to/rDataSet")
# or with
install.packages("/path/to/rDataSet", repo = NULL, type = "source")
```

## Overview

rDataSet introduces a `dataset` S3 class that extends data frames with designated ID columns. Set operations work on a cell-by-cell basis:

- **ID columns** uniquely identify each row (one-to-one relation)
- **Value presence** is determined by `is.na()` - a value is present if not `NA`
- **Set operations** compare cells at the same (row_id, col_id) position
- **Empty rows are dropped** - rows containing only `NA` values in all value columns are considered non-existent and are automatically removed during dataset construction

## Set Operations

Each set operation works in two stages:

1. **Row/Column Alignment**: Determines which rows and columns appear in the result
2. **Value Operation**: Decides which values to keep for each cell

| Operation | Rows in Result | Columns in Result | Value Operation |
|-----------|----------------|-------------------|-----------------|
| Difference (`A - B`) | All rows from A | All value columns from A | Keep values from A where B has `NA` |
| Intersection (`A > B`) | Only rows present in **both** A and B (inner join on IDs) | Only common value columns | Keep values from A where B has non-`NA` |
| Union (`A + B`) | All rows from both A and B (full join on IDs) | All value columns from both | Use A's value; fall back to B where A has `NA` |
| Equality (`A == B`) | Must match exactly (same rows) | Must match exactly (same columns) | `TRUE` if equal, `FALSE` if different, `NA` if both missing |

**Note**: ID columns must match between datasets for all operations. Value columns that exist in only one dataset are handled according to the join type (dropped, kept, or error).

## Example

```{r example}
library(rDataSet)
library(tibble)
library(dplyr)

# Create datasets with 'i' as the ID column
# Note: Rows with all-NA values in value columns are automatically dropped
A <- dataset_build(
  tibble(i = 1:10, b = if_else(1:10 %% 2 == 0, NA, 1:10), c = 1:10),
  ids = "i"
)

B <- dataset_build(
  tibble(i = 1:10, b = na_if(1:10, 3), c = 11:20),
  ids = "i"
)

# Dataset C has fewer rows (1:5) and an extra column 'd'
C <- dataset_build(
  tibble(i = 1:5, b = 10:6, d = 2),
  ids = "i"
)

# View the datasets
A
B
C
```

### Set Difference

Keep values from A where B has `NA` (uses left join, so all rows from A):

```{r difference}
A - B
```

Difference with C shows how non-common columns are dropped:

```{r difference-c}
A - C  # Only column 'b' remains (common), row ids 1:5 from A
```

### Set Intersection

Keep values from A where B has non-`NA` values (inner join on rows, common columns only):

```{r intersection}
A > B
```

Intersection with C - note only rows 1:5 and column 'b' are kept:

```{r intersection-c}
A > C
```

### Set Union

Combine values from both datasets (full join on rows, all columns):

```{r union}
A + B
```

Union with C - includes all rows (1:10) and all columns (b, c, d):

```{r union-c}
A + C
```

### Set Equality

Compare datasets cell-by-cell (requires matching rows and columns):

```{r equality}
A == A  # All TRUE (or NA for missing values)
A == B  # Shows which cells match
```

Equality fails with mismatched rows or columns:

```{r equality-error, error = TRUE}
A == C  # Error: rows and columns don't match
```

## Helper Functions

```{r helpers}
# Extract ID columns
ids(A)

# Extract value columns (non-ID columns)
vals(A)
```

## Mathematical Properties

The set operations satisfy the fundamental identity: `a = (a ∩ b) ∪ (a \ b)`

```{r identity}
# Verify: A == (A > B) + (A - B)
all(vals(A == (A > B) + (A - B)), na.rm = TRUE)
```

## License

GPL (>= 3)
